---
title: "K-Means and K-Medoids Clustering on Simulated Data"
author: "Ola Adefila"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Introduction

In this analysis, we simulate data to apply both **K-Means** and **K-Medoids** clustering algorithms, compare their results, and visualize the clustering outcomes.

We also measure the time taken for each clustering method.

## Load Required Libraries

```{r setup, message=FALSE, warning=FALSE}
# Install libraries if not already installed
# install.packages("cluster")
# install.packages("factoextra")
# install.packages("ggplot2")
# install.packages("tictoc")

# Load necessary libraries
library(cluster)      # For K-Medoids (PAM)
library(factoextra)   # For visualization
library(ggplot2)      # For plotting
library(tictoc)       # For measuring execution time
```


```{r data prep}
# Set random seed for reproducibility
set.seed(42)

# Create 3 clusters with normal distribution
n <- 6000  # Total number of data points
cluster1 <- data.frame(x = rnorm(n/3, mean = 2, sd = 0.5), y = rnorm(n/3, mean = 2, sd = 0.5))
cluster2 <- data.frame(x = rnorm(n/3, mean = 6, sd = 0.5), y = rnorm(n/3, mean = 6, sd = 0.5))
cluster3 <- data.frame(x = rnorm(n/3, mean = 10, sd = 0.5), y = rnorm(n/3, mean = 2, sd = 0.5))

# Combine clusters into one dataset
data <- rbind(cluster1, cluster2, cluster3)
colnames(data) <- c("feature1", "feature2")

# Normalize the data
scaled_data <- scale(data)

```

# Apply K-Means Clustering

```{r k-means}
tic("K-Means")  # Start timer
kmeans_result <- kmeans(scaled_data, centers = 3, nstart = 10)
toc()  # Stop timer
```

# Apply K-Medoids Clustering
```{r k-medoids}
tic("K-Medoids")  # Start timer
kmedoids_result <- pam(scaled_data, 3)
toc()  # Stop timer

```


# Prep data for visualization

```{r get memberships }
data$KMeans_Cluster <- as.factor(kmeans_result$cluster)
data$KMedoids_Cluster <- as.factor(kmedoids_result$clustering)


# Add true class labels
data$true_label <- as.factor(c(rep("1", n/3), rep("2", n/3), rep("3", n/3)))
```


# Plotting k-means clusters
```{r vis kmeans}
p1 <- ggplot(data, aes(x = feature1, y = feature2, color = KMeans_Cluster)) +
  geom_point(size = 3) +
  ggtitle("K-Means Clustering") +
  theme_minimal()

# Print the K-Means plot
print(p1)
```

# Plotting k-medoids clusters
```{r kmedoids}
p2 <- ggplot(data, aes(x = feature1, y = feature2, color = KMedoids_Cluster)) +
  geom_point(size = 3) +
  ggtitle("K-Medoids Clustering") +
  theme_minimal()

# Print the K-Medoids plot
print(p2)
```




```{r kmedoids vis}
# Mark medoids with black edges
data$is_medoid <- FALSE
data$is_medoid[kmedoids_result$id.med] <- TRUE

p3 <- ggplot(data, aes(x = feature1, y = feature2, 
                       color = KMedoids_Cluster, shape = true_label)) +
  geom_point(size = 3, alpha = 0.7) +
  geom_point(data = data[data$is_medoid, ], 
             color = "black", size = 5, shape = 1, stroke = 2) +  # Circle outline
  scale_shape_manual(values = c(16, 17, 15)) +
  ggtitle("K-Medoids Clustering") +
  labs(subtitle = "Color = Cluster, Shape = True Label, Black Circle = Medoids",
       color = "K-Medoids Cluster",
       shape = "True Class") +
  theme_minimal()

print(p3)


```



```{r pressure, echo=FALSE}
plot(pressure)
```


